Optimizing Abstract$^{\text 2}$ Machines

### 2012-11-11 paper/paper.tex
Some related work and intro.

The \emph{abstracting abstract machines}
approach~\cite{dvanhorn:VanHorn2011Abstracting,dvanhorn:VanHorn2012Systematic}
to deriving program analyses provides a straightforward way of
transforming a programming language semantics in the form of an
abstract machine, into a family of abstract interpreters parameterized
over policies for regulating the analytic precision of the control,
environment, store, and base value domains.  By taking a
machine-oriented view of computation, it becomes possible to design,
verify, and implement program analyzers for realistic language
features typically considered difficult to model.  The approach was
originally applied to features such as higher-order functions,
stack-inspection, exceptions, laziness, first-class continuations, and
garbage collection.  It has since been used to verify actor-
\cite{local:DOsualdo:12A} and
thread-based~\cite{dvanhorn:Might2011Family} parallelism and
behavioral contracts~\cite{dvanhorn:TobinHochstadt2012Higherorder}; it
has been used to model Dalvik, Erlang~\cite{local:DOsualdo:12B},
JavaScript~\cite{local:DBLP:journals/corr/abs-1109-4467},
Racket~\cite{dvanhorn:TobinHochstadt2012Higherorder}, and
Coq~\cite{local:harvard}.

The primary strength of the approach is that abstract interpreters can
be easily derived through a small number of steps from existing
machine models. % such as the CESK machine or the Lazy Krivine machine.
Moreover, since the relationships between abstract machines and
higher-level semantic models---such as definitional
interpreters~\cite{dvanhorn:reynolds-hosc98}, structured operational
semantics~\cite{dvanhorn:Plotkin1981Structural}, and reduction
semantics~\cite{dvanhorn:Felleisen2009Semantics}---are well
understood~\cite{dvanhorn:Danvy:DSc}, it is possible to navigate from
these semantic models to sound program analyzers.

@@

@@

Notation: concrete examples of programs under analysis is given in
(monochrome) Scheme notation.  Code is given in (syntax colored)
Racket.

@@

\begin{figure*}
\begin{align*}
\mathit{eval}(\mexp) &= \{ \mstate\ |\ \ev[^{\mtcntr}]{\mexp,\varnothing,\varnothing,\kmt} \multimachstep \mstate \} \text{ where }
\\[2mm]
%% EVAL
\ev{\svar\mvar,\menv,\msto,\mkont} &\machstep
\co{\mkont,\mval,\msto}
\text{ where }\mval \in \msto(\menv(\mvar))
\\
\ev{\slit\mlit,\menv,\msto,\mkont} &\machstep
\co{\mkont,\mlit,\msto}
\\
\ev{\slam\mvar\mexp,\menv,\msto,\mkont} &\machstep
\co{\mkont,\clos{\mvar,\mexp,\menv},\msto}
\\
\ev[^\mcntr]{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &\machstep
\ev[^\mcntr]{\mexpi{0},\menv,\msto',\kar[_\mlab^\mcntr]{\mexpi{1},\menv,\maddr}}
\text{ where }\maddr,\msto' = \mathit{push}^\mcntr_\mlab(\msto,\mkont)
\\
\ev[^\mcntr]{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2},\menv,\msto,\mkont} &\machstep
\ev[^\mcntr]{\mexpi0,\menv,\msto',\kif[^\mcntr]{\mexpi1,\mexpi2,\menv,\maddr}}
\text{ where }\maddr,\msto' = \mathit{push}_\mlab^\mcntr(\msto,\mkont)
\\[2mm]
%% CONTINUE
\co{\kmt,\mval,\msto} &\machstep
\ans{\msto,\mval}
\\
\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr},\mval,\msto} & \machstep
\ev[^\mcntr]{\mexp,\menv,\msto,\kfn[^\mcntr_\mlab]{\mval,\maddr}}
\\
\co{\kfn[^\mcntr_\mlab]{{\mvalx{u}},\maddr},\mval,\msto} & \machstep
\ap[^\mcntr_\mlab]{\mvalx{u},\mval,\mkont,\msto}
\text{ where }\mkont \in \msto(\maddr)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr},\strue,\msto} & \machstep
\ev[^\mcntr]{\mexpi0,\menv,\msto,\mkont}
\text{ where }\mkont\in\msto(\maddr)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr},\sfalse,\msto} & \machstep
\ev[^\mcntr]{\mexpi1,\menv,\msto,\mkont}
\text{ where }\mkont\in\msto(\maddr)
\\[2mm]
%% APPLY
\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\msto,\mkont} & \machstep
\ev[^{\mcntr'}\!]{\mexp,\menv',\msto',\mkont}
\text{ where }\menv',\msto',\mcntr' = \mathit{bind\,}^{ \mcntr}_\mlab(\menv,\msto,\mvar,\mval)
\\
\ap[^\mcntr_\mlab]{\mop,\mval,\msto,\mkont} & \machstep
\co{\mkont,\mval',\msto}
\text{ where } \mval'\in\interpdelta(\mop,\mval)
\end{align*}
\caption{Abstract$^2$ machine for ISWIM}
\label{fig:aam}
\end{figure*}

@@

\begin{figure}
\begin{alltt}
  ;; State \(\to\) Set[State]
  (define (step \(\mstate\))
    (match \(\mstate\)
      [(ev \(\mexp\) \(\rho\) \(\msto\) \(\mkont\))
       (match \(\mexp\)
         [(var\(\superscript\mlab\) \(\mvar\))
          (for/set ((\(\mval\) (lookup \(\rho\) \(\msto\) \(\mvar\))))
            (co \(\msto\) \(\mkont\) \(\mval\)))]
         [(lit\(\superscript\mlab\) \(\mlit\)) (set (co \(\msto\) \(\mkont\) \(\mlit\)))]
         [(lam\(\superscript\mlab\) \(\mvar\) \(\mexp\)) (set (co \(\msto\) \(\mkont\) (clos \(\mvar\) \(\mexp\) \(\rho\))))]
         [(app\(\superscript\mlab\) \(\mexp\subscript0\) \(\mexp\subscript1\))
          (define-values (\(\msto'\) \(\maddr\)) (push \(\mstate\)))
          (set (ev \(\mexp\subscript0\) \(\rho\) \(\msto'\) (ar \(\mexp\subscript1\) \(\rho\) \(\maddr\))))]
         [(ife\(\superscript\mlab\) \(\mexp\subscript0\) \(\mexp\subscript1\) \(\mexp\subscript2\))
          (define-values (\(\msto'\) \(\maddr\)) (push \(\mstate\)))
          (set (ev \(\mexp\subscript0\) \(\rho\) \(\msto'\) (ifk \(\mexp\subscript1\) \(\mexp\subscript2\) \(\rho\) \(\maddr\))))])]
      [(co \(\msto\) \(\mkont\) \(\mval\))
       (match \(\mkont\)
         ['mt (set (ans \(\msto\) \(\mval\)))]
         [(ar\(\superscript\mlab\) \(\mexp\) \(\rho\)) (set (ev \(\mexp\) \(\rho\) \(\msto\) (fn \(\mval\) l)))]
         [(fn\(\superscript\mlab\) \(\mval'\))
          (for/set ((\(\mkont\) (get-cont \(\msto\) l)))
            (ap \(\msto\) \(\mval'\) \(\mval\) \(\mkont\)))]
         [(fi\(\superscript\mlab\) c a \(\rho\))
          (for/set ((k (get-cont \(\msto\) l)))
            (ev (if v c \(\maddr\)) \(\rho\) \(\msto\) \(\mkont\)))])]
      [(ap \(\msto\) fun \(\maddr\) \(\mkont\))
       (match fun
         [(clos l \(\mvar\) \(\mexp\) \(\rho\))
          (define-values (\(\rho'\) \(\msto'\)) (bind \(\mstate\)))
          (set (ev \(\msto'\) \(\mexp\) \(\rho'\) \(\mkont\)))]
         [(op \(\mop\))
          (for*/set ((\(\mkont\) (get-cont \(\msto\) l))
                     (\(\mval\) (\(\interpdelta\) \(\mop\) \(\mval\))))
            (co \(\msto\) \(\mkont\) \(\mval\)))]
         [_ (set)]))]))
\end{alltt}
\caption{Implementation of machine transition relation.}
\end{figure}

@@

%% \begin{alltt}
%% (define \church2 (\(\lambda\) (f) (\(\lambda\) (x) (f (f x)))))

%% (define pred
%%   (\(\lambda\) (n)
%%     (\(\lambda\) (rf)
%%       (\(\lambda\) (rx)
%%         (((n (\(\lambda\) (g) (\(\lambda\) (h) (h (g rf)))))
%%           (\(\lambda\) (i) rx))
%%          (\(\lambda\) (id) id))))))
%% \end{alltt}

%% (define \church0 (\(\lambda\) (f0) (\(\lambda\) (x0) x0)))
%% (define \church1 (\(\lambda\) (f1) (\(\lambda\) (x1) (f1 x1))))
%% (define \church2 (\(\lambda\) (f2) (\(\lambda\) (x2) (f2 (f2 x2)))))
%% (define \church3 (\(\lambda\) (f3) (\(\lambda\) (x3) (f3 (f3 (f3 x3))))))

\begin{figure}
\begin{alltt}
(define ((((plus p1) p2) pf) x)
  ((p1 pf) ((p2 pf) x)))

(define (((mult m1) m2) mf)
  (m2 (m1 mf)))

@@

(define (zero? n)
  ((n (\(\lambda\) (zx) #f)) #t))

@@

;; multiplication distributes over addition
((church=? ((mult \church2) ((plus \church1) \church3)))
 ((plus ((mult \church2) \church1)) ((mult \church2) \church3)))
\end{alltt}
\caption{Church-numeral calculation}
\end{figure}

Where \syntax{church=?} is an equality function for Church numerals
written in terms of recursion and an iterative \syntax{pred} function.
The \syntax{plus} and \syntax{mult} functions are written as usual.

This program reaches $N$ states under concrete interpretation.

@@

We compute the semantics of a program by iterating the state
transition relation until a fixed point in the reachable states is
reached. We begin with a generic fixpoint combinator, ${\mathcal F}$:

\begin{align*}
{\mathcal F}(f)(X) &= {\mathcal G}(f, X, \varnothing) \\
{\mathcal G}(f, \varnothing, S) &= S \\
{\mathcal G}(f, X, S) &= {\mathcal G}(f, f(X) \setminus S, S \cup f(X)) \\
\end{align*}

The $\machstep$ relation is most naturally written as a function from
a state to a set of all next states. We want to find a fixpoint of
$\machstep$ on the state space, however, so we must lift it to work on
sets of states. The lifting is a natural one:

\begin{align*}
lift(f)(X) &= \bigcup\{f(x) \mid x \in X\}
\end{align*}

Then $eval(e) = {\mathcal F}(lift(\machstep))(\{\ev[^\mtcntr]{e, \varnothing, \varnothing, \kmt}\})$.

If we run this on our Church numeral example, we must be
patient. Indeed, the state space is finite, but it is exponential in
the size of the program. With $k = 0$, the store can be in one of ${\mathcal
  O}(2^n)$ many configurations! We take a step toward a polynomial
complexity baseline by lifting the store out so that there is one
shared amongst all states.

@@

A common technique to faster convergence in flow analyses is to share
a common store that is monotonically updated as interpretation
continues. In the ISWIM language with only unary operators, this makes
our 0-CFA quartic in the size of the program. Not great, but
polynomial.

We can reuse the reduction relation we have defined, but with some
scaffolding around it. A state in the store-widened machine is a pair
of a set of ISWIM states without a store, and a store. A step of this
machine will pair the store with each of the states in this set,
reduce, and combine the output stores to create the next. Let $wn :
\widehat{State} \times Store \to State$ be the metafunction that adds
a store to a storeless state, $nw$ removes the store, and $c.\msto$
projects the store out of a state $c$. Formally,

\begin{align*}
(cs, \msto) &\widehat{\machstep} (cs', \msto') \\
\mbox{where } CS' &= lift(\machstep)(\{ wn(c, \msto) \mid c \in cs\}) \\
              cs' &= \{nw(C') \mid C' \in CS'\} \\
              \msto' &= \bigcup\{C'.\msto \mid C' \in CS'\}
\end{align*}

@@

The final approximation we make to get to our baseline analysis is
store-allocating results of application sub-expressions. Notice that
the $\kfn{}$ continuation stores a value - this makes the space of
continuations quadratic rather than linear in 0-CFA. We instead
allocate an address for the value position when we create the
continuation. This address and the tail address are both determined by
the label of the application point, so the space becomes linear and
the overall complexity drops to cubic. Note that this is a critical
abstraction in languages with n-ary functions, since otherwise the
continuation space becomes ${\mathcal O}(n^n)$!

\begin{align*}
\ev{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &\machstep
\ev{\mexpi{0},\menv,\msto\sqcup[\mlab \mapsto \{\mkont\}],\kar{\mexpi{1},\mlab}}
\\
\co{\kar{\mexp,\menv,\mlab},\mval,\msto} & \machstep
\ev{\mexp,\menv,\msto\sqcup[\mlab^f \mapsto \{\mval\}],\kfn{\mlab^f,\mlab}}
\\
\co{\kfn{\mlab^f,\mlab},\mval,\msto} & \machstep
\ap[_\mlab]{\mvalx{u},\mlab^a,\mkont,\msto\sqcup[\mlab^a \mapsto \{\mval\}]}
\\
\text{ where } &\mkont \in \msto(\maddr), \mvalx{u} \in \msto(\maddr^f)
\\
\ap[_\mlab]{\clos{\mvar,\mexp,\menv},\mlab^a, \msto,\mkont} & \machstep
\ev{\mexp,\menv[\mvar\mapsto\mvar],\msto\sqcup[\mvar \mapsto \msto(\mlab^a)],\mkont}
\\
\ap[_\mlab]{\mop,\mlab^a,\msto,\mkont} & \machstep
\co{\mkont,\mval',\msto}
\\
\text{ where }& \mval'\in \{\interpdelta(\mop,\mval) \mid \mval \in \msto(\mlab^a)\}
\end{align*}

This extra store-allocation is effectively naming all intermediate
results, and thus the precision aligns with an analysis specialized to
either ANF or CPS.

@@

The first problem that we realized with the AAM approach was that much
of the non-determinism is spurious. For example, in a function
application,
\begin{alltt}
(f x y)
\end{alltt}

where all variables have been joined with multiple values, each
variable reference transitions to several new states that comprise the
cartesian product of all the values of {\tt f}, {\tt x} and {\tt y},
only to be joined back together in the application position. We would see transitions of the following shape:
\begin{center}
\includegraphics[scale=0.3]{fanout.pdf}
\end{center}
For this reason, we {\it delay} the non-determinism
until it is needed in {\it strict contexts} (such as the guard for an
{\tt if} a called procedure, or a numerical primitive application). We instead get transitions of this shape:
\begin{center}
\includegraphics[scale=0.3]{lazy.pdf}
\end{center}

@@

\begin{align*}
force(\msto,\saddr\maddralt) &= \msto(\maddralt)\\
force(\msto,\mval) &= \{\mval\}\\
\ev{\svar{\mvar},\menv,\mkont,\msto} &\machstep\;
\co{\mkont,\saddr{\menv(\mvar)},\msto} \\
\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr},\mval,\msto}
&\machstep\;
\ev[^\mcntr]{\mexp,\menv,\msto\sqcup[\maddr^f \mapsto force(\msto,\mval)],\kfn[^\mcntr_\mlab]{\maddr^f,\maddr}} \\
bind^\mcntr_\mlab(\menv,\msto,\mvar,\mval) &= \menv[\mvar \mapsto
  \maddr], \msto\sqcup[\maddr \mapsto force(\msto, \mval)], \mcntr' \\
\mbox{where } \mcntr' &= \lfloor\mlab\mcntr\rfloor_k \\
              \maddr &= \mvar\mcntr'
\end{align*}

@@

The {\tt do} form has the special job of maintaining the current store
and accumulated states (if indeed those are being passed around). It
uses Racket's syntax parameters ~\cite{ianjohnson:eli/stxparam} to
track the most recent bindings of these accumulators so that {\tt
  yield-meaning} can refer to them (say {\tt $\sigma$p} and {\tt
  statep}). In the case of state passing and set-accumulated states, its
definition is simply

@@

@@

@@

\appendix
\section{Relation to Uniform \(k\)-CFA (A Case Against Acceptability)}

>>>>

The \emph{abstracting abstract machines}
approach~\cite{dvanhorn:VanHorn2011Abstracting,dvanhorn:VanHorn2012Systematic}
to deriving program analyses provides a straightforward way of
transforming a programming language semantics in the form of an
abstract machine, into a family of abstract interpreters parameterized
over policies for regulating the analytic precision of the control,
environment, store, and base value domains.  By taking a
machine-oriented view of computation, it becomes possible to design,
verify, and implement program analyzers for realistic language
features typically considered difficult to model.  The approach was
originally applied to features such as higher-order functions,
stack-inspection, exceptions, laziness, first-class continuations, and
garbage collection.  It has since been used to verify actor-
\cite{local:DOsualdo:12A} and
thread-based~\cite{dvanhorn:Might2011Family} parallelism and
behavioral contracts~\cite{dvanhorn:TobinHochstadt2012Higherorder}; it
has been used to model Coq~\cite{local:harvard},
Dalvik~\cite{local:dalvik}, Erlang~\cite{local:DOsualdo:12B},
JavaScript~\cite{local:DBLP:journals/corr/abs-1109-4467}, and
Racket~\cite{dvanhorn:TobinHochstadt2012Higherorder}.

The primary strength of the approach is that abstract interpreters can
be easily derived through a small number of steps from existing
machine models.  Since the relationships between abstract machines and
higher-level semantic models---such as definitional
interpreters~\cite{dvanhorn:reynolds-hosc98}, structured operational
semantics~\cite{dvanhorn:Plotkin1981Structural}, and reduction
semantics~\cite{dvanhorn:Felleisen2009Semantics}---are well
understood~\cite{dvanhorn:Danvy:DSc}, it is possible to navigate from
these high-level semantic models to sound program analyzers in a
systematic way.  Moreover, since these analyses so closely resemble a
language's interpreter (a) implementing an analysis requires little
more than implementing an interpreter, (b) a single implementation can
serve as both an interpreter and analyzer, and (c) verifying the
correctness of the implementation is straightforward.

However, there is a considerable weakness with the approach: an
analyzer designed and implemented by following the AAM recipe is
prohibitively inefficient without both further approximation and
further implementation effort.

In this article, we develop a systematic approach to deriving the
feasible implementation of an abstract machine based analyzer.

@@

\includegraphics[width=2.5in]{rel-time-church.ps}

@@

%% Notation: concrete examples of programs under analysis is given in
%% (monochrome) Scheme notation.  Code is given in (syntax colored)
%% Racket.

@@

\begin{figure*}
\begin{align*}
\mathit{eval}(\mexp) &= \{ \mstate\ |\ \ev[^{\mtcntr}]{\mexp,\varnothing,\varnothing,\kmt} \multimachstep \mstate \} \text{ where }
\\[2mm]
%% EVAL
\ev{\svar\mvar,\menv,\msto,\mkont} &\machstep
\co{\mkont,\mval,\msto}
\text{ where }\mval \in \msto(\menv(\mvar))
\\
\ev{\slit\mlit,\menv,\msto,\mkont} &\machstep
\co{\mkont,\mlit,\msto}
\\
\ev{\slam\mvar\mexp,\menv,\msto,\mkont} &\machstep
\co{\mkont,\clos{\mvar,\mexp,\menv},\msto}
\\
\ev[^\mcntr]{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &\machstep
\ev[^\mcntr]{\mexpi{0},\menv,\msto',\kar[_\mlab^\mcntr]{\mexpi{1},\menv,\maddr}}
\text{ where }\maddr,\msto' = \mathit{push}^\mcntr_\mlab(\msto,\mkont)
\\
\ev[^\mcntr]{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2},\menv,\msto,\mkont} &\machstep
\ev[^\mcntr]{\mexpi0,\menv,\msto',\kif[^\mcntr]{\mexpi1,\mexpi2,\menv,\maddr}}
\text{ where }\maddr,\msto' = \mathit{push}_\mlab^\mcntr(\msto,\mkont)
\\[2mm]
%% CONTINUE
\co{\kmt,\mval,\msto} &\machstep
\ans{\msto,\mval}
\\
\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr},\mval,\msto} & \machstep
\ev[^\mcntr]{\mexp,\menv,\msto,\kfn[^\mcntr_\mlab]{\mval,\maddr}}
\\
\co{\kfn[^\mcntr_\mlab]{{\mvalx{u}},\maddr},\mval,\msto} & \machstep
\ap[^\mcntr_\mlab]{\mval,\mvalx{u},\mkont,\msto}
\text{ where }\mkont \in \msto(\maddr)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr},\strue,\msto} & \machstep
\ev[^\mcntr]{\mexpi0,\menv,\msto,\mkont}
\text{ where }\mkont\in\msto(\maddr)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr},\sfalse,\msto} & \machstep
\ev[^\mcntr]{\mexpi1,\menv,\msto,\mkont}
\text{ where }\mkont\in\msto(\maddr)
\\[2mm]
%% APPLY
\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\msto,\mkont} & \machstep
\ev[^{\mcntr'}\!]{\mexp,\menv',\msto',\mkont}
\text{ where }\menv',\msto',\mcntr' = \mathit{bind\,}^{ \mcntr}_\mlab(\menv,\msto,\mvar,\mval)
\\
\ap[^\mcntr_\mlab]{\mop,\mval,\msto,\mkont} & \machstep
\co{\mkont,\mval',\msto}
\text{ where } \mval'\in\interpdelta(\mop,\mval)
\end{align*}
\caption{Abstract$^2$ machine for ISWIM}
\label{fig:aam}
\end{figure*}

@@

\begin{figure}
\begin{alltt}
  ;; State \(\to\) \(\Set(\)State\()\)
  (define (step \(\mstate\))
    (match \(\mstate\)
      [(ev \(\mexp\) \(\rho\) \(\msto\) \(\mkont\))
       (match \(\mexp\)
         [(var\(\superscript\mlab\) \(\mvar\))
          (for/set ((\(\mval\) (lookup \(\rho\) \(\msto\) \(\mvar\))))
            (co \(\msto\) \(\mkont\) \(\mval\)))]
         [(lit\(\superscript\mlab\) \(\mlit\)) (set (co \(\msto\) \(\mkont\) \(\mlit\)))]
         [(lam\(\superscript\mlab\) \(\mvar\) \(\mexp\)) (set (co \(\msto\) \(\mkont\) (clos \(\mvar\) \(\mexp\) \(\rho\))))]
         [(app\(\superscript\mlab\) \(\mexp\subscript0\) \(\mexp\subscript1\))
          (define-values (\(\msto'\) \(\maddr\)) (push \(\mstate\)))
          (set (ev \(\mexp\subscript0\) \(\rho\) \(\msto'\) (ar \(\mexp\subscript1\) \(\rho\) \(\maddr\))))]
         [(ife\(\superscript\mlab\) \(\mexp\subscript0\) \(\mexp\subscript1\) \(\mexp\subscript2\))
          (define-values (\(\msto'\) \(\maddr\)) (push \(\mstate\)))
          (set (ev \(\mexp\subscript0\) \(\rho\) \(\msto'\) (ifk \(\mexp\subscript1\) \(\mexp\subscript2\) \(\rho\) \(\maddr\))))])]
      [(co \(\msto\) \(\mkont\) \(\mval\))
       (match \(\mkont\)
         ['mt (set (ans \(\msto\) \(\mval\)))]
         [(ar\(\superscript\mlab\) \(\mexp\) \(\rho\)) (set (ev \(\mexp\) \(\rho\) \(\msto\) (fn \(\mval\) l)))]
         [(fn\(\superscript\mlab\) \(\mval'\))
          (for/set ((\(\mkont\) (get-cont \(\msto\) l)))
            (ap \(\msto\) \(\mval'\) \(\mval\) \(\mkont\)))]
         [(fi\(\superscript\mlab\) c a \(\rho\))
          (for/set ((k (get-cont \(\msto\) l)))
            (ev (if v c \(\maddr\)) \(\rho\) \(\msto\) \(\mkont\)))])]
      [(ap \(\msto\) fun \(\maddr\) \(\mkont\))
       (match fun
         [(clos l \(\mvar\) \(\mexp\) \(\rho\))
          (define-values (\(\rho'\) \(\msto'\)) (bind \(\mstate\)))
          (set (ev \(\msto'\) \(\mexp\) \(\rho'\) \(\mkont\)))]
         [(op \(\mop\))
          (for*/set ((\(\mkont\) (get-cont \(\msto\) l))
                     (\(\mval\) (\(\interpdelta\) \(\mop\) \(\mval\))))
            (co \(\msto\) \(\mkont\) \(\mval\)))]
         [_ (set)]))]))
\end{alltt}
\caption{Implementation of machine transition relation.}
\end{figure}

@@

\begin{figure}
\begin{alltt}
(define (zero? n) ((n (\(\lambda\) (zx) #f)) #t))
(define (((mult m1) m2) mf) (m2 (m1 mf)))
(define ((((plus p1) p2) pf) x)
  ((p1 pf) ((p2 pf) x)))

@@

@@

;; multiplication distributes over addition
((church=? ((mult \church2) ((plus \church1) \church3)))
 ((plus ((mult \church2) \church1)) ((mult \church2) \church3)))
\end{alltt}
\caption{Church-numeral calculation}
\label{fig:church}
\end{figure}

Figure~\ref{fig:church} gives a small example of computing with Church
numerals.  The example, while small, is informative: (1) it can be
written in almost any modern programming language\footnote{It is
  written here using Scheme notation and makes use of the shorthands:
%
\begin{alltt}
   (define (((f x) y) z) ...) \(\equiv\)

   (define f (\(\lambda\) (x) (\(\lambda\) (y) (\(\lambda\) (z) ...))))
\end{alltt}

\begin{alltt}
   \church{n} \(\equiv\) (\(\lambda\) (s) (\(\lambda\) (z) (s\(\subscript1\) (s\(\subscript2\) ... (s\(\subscript{n}\) z)))))
\end{alltt}
},
%
and (2) it stresses an analyzer's ability to deal with complicated
environment and control structure arising from the use of higher-order
functions to model arithmetic.

Our baseline analyzer takes NN:NN minutes to analyze this program
(without store widening it times out after 30 minutes).  This is on
par with the performance of Sergey,~et al.'s analyzer, which takes
NN:NN minutes.

@@

\begin{alltt}
;; Expr \(\to\) \(\Set(\)State\()\)
(define (eval e)
  (fix step (set (inj e))))
\end{alltt}

We compute the semantics of a program by iterating the state
transition relation until a fixed point in the reachable states is
reached.

\begin{alltt}
;; \(\forall\) (X) ((X \(\to\) \(\Set(\)X\()\)) \(\to\) (\(\Set(\)X\()\) \(\to\) \(\Set(\)X\()\)))
(define ((appl f) s)
  (for/fold ([i (set)])
    ([x (in-set s)])
    (\(\cup\) i (f x))))

;; Calculate fixpoint of (appl f).
;; \(\forall\) (X) ((X \(\to\) Set[X]) Set[X] \(\to\) Set[X])
(define (fix f s)
  (let loop ((accum (set)) (front s))
    (if (set-empty? front)
        accum
        (let ((new-front ((appl f) front)))
          (loop (\(\cup\) accum front)
                (new-front . \(\setminus\) . accum))))))
\end{alltt}

@@

\begin{alltt}
;; \(\widehat{State}\) = (cons (Set Conf) Store)

;; (State \(\to\) \(\Set(\)State\()\)) \(\to\) \(\widehat{State}\) \(\to\) \(\widehat{State}\)
(define ((wide-step step) state)
  (match state
    [(cons cs \(\msto\))
     (define ss ((appl step)
                 (for/set ([c cs]) (c->s c \(\msto\)))))
     (set (cons (for/set ([s ss]) (s->c s))
                (\(\sqcup\)-stores ss)))]))
\end{alltt}

@@

@@

The first problem that we realized with the AAM approach was that much
of the non-determinism is spurious. For example, in a function
application,
\begin{alltt}
(f x y z w)
\end{alltt}
where all variables have been joined with multiple values, each
variable reference transitions to several new states that comprise the
cartesian product of all the values of {\tt x}, {\tt y}, {\tt z} and
{\tt w}, only to be joined back together in {\tt f}'s bindings. For
this reason, we {\it delay} the non-determinism until it is needed in
{\it strict contexts} (such as the guard for an {\tt if} a called
procedure, or a numerical primitive application).

@@

\begin{align*}
force(\msto,\clos{\mvar,\mexp,\menv}) &= \{\clos{\mvar,\mexp,\menv}\}\\
force(\msto,\saddr\maddralt) &= \msto(\maddralt)\\
\ev{\svar{\mvar},\menv,\mkont,\msto} &\machstep\;
\co{\mkont,\saddr{\menv(\mvar)},\msto} \\
\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr},v,\msto}
&\machstep\;
\ev[^\mcntr]{\mexp,\menv,\msto,\kfn[^\mcntr_\mlab]{\mval,\maddr}} \\
\mbox{where } &\mval' \in force(\msto,\mval) \\
bind^\mcntr_\mlab(\menv,\msto,\mvar,\mval) &= \menv[\mvar \mapsto
  \maddr], \msto\sqcup[\maddr \mapsto force(\msto, \mval)], \mcntr' \\
\mbox{where } \mcntr' &= \lfloor\mlab\mcntr\rfloor_k \\
              \maddr &= \mvar\mcntr'
\end{align*}

The force for the second rule is necessary to keep the same type for
the \(\kfn{}\) continuation, but we would rather use an address. This
introduces (acceptably) more imprecision, but has the same precision as an ANF
analysis\footnote{Exercise for the reader.}.

@@

The {\tt do} form has the special job of maintaining the current store
and accumulated states (if indeed those are being passed around). It
uses Racket's syntax parameters ~\cite{dvanhorn:eli/stxparam} to
track the most recent bindings of these accumulators so that {\tt
  yield-meaning} can refer to them (say {\tt $\sigma$p} and {\tt
  statep}). In the case of state passing and set-accumulated states, its
definition is simply

@@

\paragraph{Constraint-based program analysis for higher-order languages}

Constraint-based program analyses
(e.g.~\cite{dvanhorn:nielson-nielson-popl97,dvanhorn:wright-jagannathan-toplas98,dvanhorn:Meunier2006Modular})
typically compute sets of abstract values for each program point.
These values approximate the values arising at run-time for each
program point.  Value sets are computed as the least solution to a set
of constraints (either inclusion or equality constraints).  The
constraints must be designed and proved as a sound approximation of
the semantics.  Efficient implementations of these kinds of analyses
often take the form of worklist-based graph algorithms for contraint
solving, and are thus quite different from the interpreter
implementation.  The approach thus requires effort in constraint
system design and implementation, and the resulting system require
verification effort to prove the constraint system is sound and that
the implementation is correct.

This effort increases substantially as the comlexity of the analyzed
language increases.  Both the work of maintaining the concrete
semantics and constraint system (and the relations between them) must
be scaled simultaneously.  However, constraint systems, which have
been extensively studied in their own right, enjoy efficient
implementation techniques and can be expressed in declarative logic
languages that are heavily
optimized~\cite{dvanhorn:bravenboer-smaragdakis-oopsla09}.
Consequently, constraint-based analyses can be computed quickly.  For
example, Jagannathan and Wright's polymorphic splitting
implementation~\cite{dvanhorn:wright-jagannathan-toplas98} analyses
the church numeral example of figure~\ref{fig:church} in 7
milliseconds, about 25 times faster than the fastest implementation
considered here.  These analyses compute very different things (as
detailed in section~\ref{sec:accept}), so the performance comparison
is not apples-to-apples.  We view the primary contribution of this
work as a systematic path that eases the design, verification, and
implementation of analyses using the abstracting abstract machine
approach to within a factor of performant constraint-based analyses.

@@

%% \acks Sam Tobin-Hochstadt for encouragement and feedback -- he was
%% the first to prompt me to look into how make effective
%% implementations of the AAM approach.

%% NSF, DARPA

@@

\appendix
\section{Relation to Uniform \(k\)-CFA (A Case Against Acceptability)}
\label{sec:accept}
